Using Occupancy Theory and Hierarchical Models to Estimate Otter Population in Southeastern Minnesota \\ {\large Senior Comprehensive Exercise: Carleton College}

### 2010-03-02 Paper.tex
Readded Sarah's edits, modified the code appendix slightly.

\appendix
\section{Deriving \(E[S]\) and \(SD[S]\)}
\label{math}
    We let \(A = \text{\# of
    0-1's}\), \(B=\text{\# of 1-1's}\) and \(p=P(``1")\text{ in the sequence}\).
    In order to find \(E[s]\), we first needed to find \(E[A]\) and
    \(E[B]\).

@@

    % E[s]
    \begin{equation*}
        \begin{split}
        E[s]& =E[A]-E[B] \\
                  & =(n-1)\left((1-p)p-p^2\right)
        \end{split}
    \end{equation*}

    The standard deviation was a little more difficult.

    % Var[A]
    \begin{equation*}
        \begin{split}
        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[I_i=(0,1),I_{i+1}=
                 (1,1)] \\
              & =(n-1)P((0,1))(1-P((0,1))) \\
              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(0,1),I_{i+1}=(0,1)]-E[I_i=
                (0,1)]E[I_{i+1}=(0,1)]\bigg) \\
              & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
                \bigg) \\
              & =(1-p)p(-1+n+5p-3np-5p^2+3np^2)
        \end{split}
    \end{equation*}

    % Var[B]
    \begin{equation*}
        \begin{split}
        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[I_i=(1,1),I_{i+1}=
                (1,1)] \\
              & =(n-1)P((1, 1))(1-P((1, 1))) \\
              & \quad +2\sum_{i=1}^{n-2}\bigg(E[I_i=(1,1),I_{i+1}=(1,1)]-E[I_i=
                (1,1)]E[I_{i+1}=(1,1)]\bigg) \\
              & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
              & =(1-p)p^2(-1+n-5p+3np)
        \end{split}
    \end{equation*}

@@

    % Var[s]
    \begin{equation*}
        \begin{split}
        Var[s]& =Var[A-B] \\
                    & =Var[A]+Var[B]-2Cov[A,B] \\
                    & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
        \end{split}
    \end{equation*}

\section{Standard hierarchical model}
\scriptsize
\begin{verbatimtab}
otterModel <- function() {
    # Initialization of variables to estimate.
    psi ~ dunif(0,1)
    theta ~ dunif(0,1)
    a ~ dgamma(1,.1)
    b ~ dgamma(1,.1)
    c ~ dgamma(1,.1)
    d ~ dgamma(1,.1)
    for (observer in 1:observers) {
        #p[observer] ~ dbeta(a,b)
        p[observer] ~ dbeta(a,b) %_% I(.5, 1)
        E[observer] ~ dbeta(c,d) %_% I(0, .5)
    }

@@

    # Super difficult stuff.  (Not really.)
    for (i in 1:numDataPts) {
        # equals(track.count[...], 0) == 1 when there is no track to observe.
        mu[i] <- 
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        # probability of false detection, given track and occupancy state
        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        prob[i] <- mu[i] + nu[i]
        #prob[i] <- mu[i]
        y[i] ~ dbern(prob[i])
    }
}

@@

modelUpdate(10000)
samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi", "theta"))
#samplesSet(c("a", "b", "p[]", "psi", "theta"))
modelUpdate(10000)
\end{verbatimtab}
%}

\section{CAR model}
\scriptsize
\begin{verbatimtab}
otterModel <- function() {
    for (site in 1:sites) {
        logit(theta[site]) <- alpha1[site] + alpha0
    }
    for (i in 1:sumNumNeigh) {
        weights[i] <- 1
    }
    alpha1[1:sites] ~ car.normal(adj[], weights[], num[], tau)
    alpha0 ~ dflat()
    tau ~ dgamma(.5,.0005)
    psi ~ dunif(0,1) 
    a ~ dgamma(1,.1)
    b ~ dgamma(1,.1)
    #c ~ dgamma(1,.1)
    #d ~ dgamma(1,.1)
    for (observer in 1:observers) {
        p[observer] ~ dbeta (a,b)
	  #p[observer] ~ dbeta(a,b) %_% I(.6, 1)
        #E[observer] ~ dbeta(c,d) %_% I(0, .3)
    }

@@

    # Super difficult stuff.  (Not really.)
    for (i in 1:numDataPts) {
        # equals(track.count[...], 0) == 1 when there is no track to observe.
        mu[i] <- 
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        # probability of false detection, given track and occupancy state
        #nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        #prob[i] <- mu[i] + nu[i]
        prob[i] <- mu[i]
	  y[i] ~ dbern(prob[i])
    }
}

>>>>

\appendix
\section{Deriving the mean and standard deviation of our \(S\) statistic}
\label{math}
    Given a binary sequence \(X_1, ..., X_n\) of independently and identically
    distributed 0's and 1's, let \(A=\text{\# of successive 0-1's}\), \(B=
    \text{\# of successive 1-1's}\) and \(p=\text{the proportion of 1's in the
    sequence}\). In order to find \(E[S]\), we first needed to find \(E[A]\) and
    \(E[B]\).

@@

    % E[S]
    \begin{equation*}
        \begin{split}
        E[S]& =E[A]-E[B] \\
            & =(n-1)\left((1-p)p-p^2\right)
        \end{split}
    \end{equation*}

    For the standard deviation,

    % Var[A]
    \begin{equation*}
        \begin{split}
        Var[A]& =\sum_{i=1}^{n-1}Var[A]+2\sum_{i=1}^{n-2}Cov[\{X_i=0, 
                X_{i+1}=1\}, \{X_{i+1}=0, X_{i+2}=1\}] \\
              & =(n-1)p(1-(1-p)p) \\
              & \quad +2\sum_{i=1}^{n-2}\bigg(E[\{X_i=0, X_{i+1}=1\},
                \{X_{i+1}=0, X_{i+2}=1\}] \\
              & \qquad\qquad\quad -E[X_i=0,X_{i+1}=1]E[X_{i+1}=0,X_{i+2}=1]
                \bigg) \\
              & =(n-1)(1-p)p(1-(1-p)p)+2\sum_{i=1}^{n-2}\bigg(0-(1-p)p(1-p)p
                \bigg) \\
              & =(1-p)p(-1+n+5p-3np-5p^2+3np^2)
        \end{split}
    \end{equation*}

    % Var[B]
    \begin{equation*}
        \begin{split}
        Var[B]& =\sum_{i=1}^{n-1}Var[B]+2\sum_{i=1}^{n-2}Cov[\{X_i=1, 
                X_{i+1}=1\}, \{X_{i+1}=1, X_{i+2}=1\}] \\
              & =(n-1)p^2(1-p^2) \\
              & \quad +2\sum_{i=1}^{n-2}\bigg(E[\{X_i=1, X_{i+1}=1\},
                \{X_{i+1}=1, X_{i+2}=1\}] \\
              & \qquad\qquad\quad -E[X_i=1, X_{i+1}=1]E[X_{i+1}=1,X_{i+2}=1]
                \bigg) \\
              & =(n-1)p^2(1-p^2)+2(n-2)(p^3-p^4) \\
              & =(1-p)p^2(-1+n-5p+3np)
        \end{split}
    \end{equation*}

@@

    % Var[S]
    \begin{equation*}
        \begin{split}
        Var[S]& =Var[A-B] \\
              & =Var[A]+Var[B]-2Cov[A,B] \\
              & =(1-p)p(-1+n+8p-4np-20p^2+12np^2)
        \end{split}
    \end{equation*}

\section{Standard hierarchical model}
\scriptsize
\begin{verbatimtab}
otterModel <- function() {
    # Initialization of variables to estimate.
    psi ~ dunif(0,1)
    theta ~ dunif(0,1)
    a ~ dgamma(1,.1)
    b ~ dgamma(1,.1)
    c ~ dgamma(1,.1)
    d ~ dgamma(1,.1)
    for (observer in 1:observers) {
        p[observer] ~ dbeta(a,b) %_% I(.5, 1)
        E[observer] ~ dbeta(c,d) %_% I(0, .5)
    }

@@

    for (i in 1:numDataPts) {
        # equals(track.count[...], 0) == 1 when there is no track to observe.
        mu[i] <- 
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        # probability of false detection, given track and occupancy state
        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        prob[i] <- mu[i] + nu[i]
        y[i] ~ dbern(prob[i])
    }
}

@@

modelUpdate(10000)
samplesSet(c("a", "b", "c", "d", "p[]", "E[]", "psi", "theta"))
#samplesSet(c("a", "b", "p[]", "psi", "theta"))
modelUpdate(10000)
\end{verbatimtab}

\section{CAR model}
\scriptsize
\begin{verbatimtab}
otterModel <- function() {
    for (site in 1:sites) {
        logit(theta[site]) <- alpha1[site] + alpha0
    }
    for (i in 1:sumNumNeigh) {
        weights[i] <- 1
    }
    alpha1[1:sites] ~ car.normal(adj[], weights[], num[], tau)
    alpha0 ~ dflat()
    tau ~ dgamma(.5,.0005)
    psi ~ dunif(0,1) 
    a ~ dgamma(1,.1)
    b ~ dgamma(1,.1)
    #c ~ dgamma(1,.1)
    #d ~ dgamma(1,.1)
    for (observer in 1:observers) {
        p[observer] ~ dbeta(a,b) %_% I(.6, 1)
        E[observer] ~ dbeta(c,d) %_% I(0, .3)
    }

@@

    for (i in 1:numDataPts) {
        # equals(track.count[...], 0) == 1 when there is no track to observe.
        mu[i] <- 
            p[obsV[i]] * (1 - equals(trackSum[siteV[i], eventV[i], dayV[i]], 0))
        # probability of false detection, given track and occupancy state
        nu[i] <- E[obsV[i]] * equals(trackSum[siteV[i], eventV[i], dayV[i]], 0)
        prob[i] <- mu[i] + nu[i]
        y[i] ~ dbern(prob[i])
    }
}
